---
title: "2MR Tutorial"
author: "Christine Le"
date: "6/8/2020"
output: html_document
---
# Reference :https://mrcieu.github.io/TwoSampleMR/index.html

## I. Basic WorkFlow
```{r}
library(TwoSampleMR)

# List available GWASs
ao <- available_outcomes()

# Get instruments
#body mass index
exposure_dat <- extract_instruments("ieu-a-2")

# Get effects of instruments on outcome
#CHD
outcome_dat <- extract_outcome_data(snps=exposure_dat$SNP, outcomes="ieu-a-7")

# Harmonise the exposure and outcome data
dat <- harmonise_data(exposure_dat, outcome_dat)

# Perform MR
res <- mr(dat)

```


## II. Exposures

### A. Importing exposure data with column names
```{r}

#file path
bmi_file <- system.file("extdata", "bmi.txt", package="TwoSampleMR")

#read file 
bmi_exp_dat <- read_exposure_data(bmi_file)
```

### B. importing exposure data without correct column names

```{r}
#file path
bmi2_file <- system.file("extdata/bmi.csv", package="TwoSampleMR")

#default to exposure if no phenotype name specified
#match up column names to match with format for 2MR
bmi_exp_dat <- read_exposure_data(
    filename = bmi2_file,
    sep = ",",
    snp_col = "rsid",
    beta_col = "effect",
    se_col = "SE",
    effect_allele_col = "a1",
    other_allele_col = "a2",
    eaf_col = "a1_freq",
    pval_col = "p-value",
    units_col = "Units",
    gene_col = "Gene",
    samplesize_col = "n"
)

#input values in exposure column to be more specific
bmi_exp_dat$exposure <- "BMI"
```


### C. Format exisiting df to work for twosampleMR
```{r}
#example df
#make sure to name columns correctly or it will not be converted over
random_df <- data.frame(
    SNP = c("rs1", "rs2"),
    beta = c(1, 2),
    se = c(1, 2),
    effect_allele = c("A", "T")
)
random_exp_dat <- format_data(random_df, type="exposure")
random_exp_dat
```


### D, Instruments that have significant associations and can be used for MR analysis


```{r}
library(MRInstruments)

#from GWAS
data(gwas_catalog)
head(gwas_catalog)

#use data from specific paper/study and phenotype
bmi_gwas <- subset(gwas_catalog, grepl("Speliotes", Author) & Phenotype == "Body mass index")
bmi_exp_dat <- format_data(bmi_gwas)

#for metabolites
data(metab_qtls)
head(metab_qtls)
# as above with metabolite data
ala_exp_dat <- format_metab_qtls(subset(metab_qtls, phenotype=="Ala"))

#for proteins
data(proteomic_qtls)
head(proteomic_qtls)

apoh_exp_dat <- format_proteomic_qtls(subset(proteomic_qtls, analyte=="ApoH"))

#for genes
data(gtex_eqtl)
head(gtex_eqtl)

irak1bp1_exp_dat <- format_gtex_eqtl(subset(gtex_eqtl, gene_name=="IRAK1BP1" & tissue == "Adipose Subcutaneous"))
#by dna methylation
data(aries_mqtl)
head(aries_mqtl)

cg25212131_exp_dat <- format_aries_mqtl(subset(aries_mqtl, cpg=="cg25212131" & age == "Birth"))

```


### E. Ensuring independence via clumping(reducing covariance)

```{r}

#will calculate the LD values bwtn SNPs and keep only the SNP with the lowest value among SNPs with LD r^2 above threshold 
bmi_exp_dat <- clump_data(bmi_exp_dat)
```

## III. Outcome Data

### A.  Extract variants from outcome trait

extract specific SNPs (prior knowledge or lit review or for SNPs examined in study)
```{r}

#get exposure data for SNP
bmi_exp_dat <- extract_instruments(outcomes='ieu-a-2')
head(bmi_exp_dat)

#search for outcome "heart disease"
ao[grepl("heart disease", ao$trait), ]

#ex : choose ieu-a-7 (e.g. most recent)

chd_out_dat <- extract_outcome_data(
    snps = bmi_exp_dat$SNP,
    outcomes = 'ieu-a-7'
)
```


### B. LDs - proxies for SNPs

extract specific SNPs (prior knowledge or lit review or only SNPs collected for particular study)
```{r}

#get exposure data for SNP
bmi_exp_dat <- extract_instruments(outcomes='ieu-a-2')
head(bmi_exp_dat)

#search for outcome "heart disease"
ao[grepl("heart disease", ao$trait), ]

#ex : choose ieu-a-7 (e.g. most recent)

chd_out_dat <- extract_outcome_data(
    snps = bmi_exp_dat$SNP,
    outcomes = 'ieu-a-7'
)
```


### C. if SNP data is not available in outcome data find LD proxies using 1000 genome data.

The parameters for handling LD proxies are as follows:

    proxies = TRUE or FALSE (TRUE by default)
    rsq = numeric value of minimum rsq to find a proxy. Default is 0.8, minimum is 0.6
    palindromes = Allow palindromic SNPs? Default is 1 (yes)
    maf_threshold = If palindromes allowed then what is the maximum minor allele frequency of   palindromes allowed  Default is 0.3.


outcome data from local GwAS data
EXAMPLE :
outcome_dat <- read_outcome_data(
    snps = bmi_exp_dat$SNP,
    filename = "gwas_summary.csv",
    sep = ",",
    snp_col = "rsid",
    beta_col = "effect",
    se_col = "SE",
    effect_allele_col = "a1",
    other_allele_col = "a2",
    eaf_col = "a1_freq",
    pval_col = "p-value",
    units_col = "Units",
    gene_col = "Gene",
    samplesize_col = "n"
)

Parameters
The extract_outcome_data function returns a table of SNP effects for the requested SNPs on the requested outcomes. The format of the data is similar to the exposure data format, except the main columns are as follows:

    SNP
    beta.outcome
    se.outcome
    samplesize.outcome
    ncase.outcome
    ncontrol.outcome
    pval.outcome
    eaf.outcome
    effect_allele.outcom
    other_allele.outcome
    units.outcome
    outcome
    consortium.outcome
    year.outcome
    pmid.outcome
    id.outcome
    originalname.outcome
    proxy.outcome
    target_snp.outcome
    proxy_snp.outcome
    target_a1.outcome
    target_a2.outcome
    proxy_a1.outcome
    proxy_a2.outcome
    mr_keep.outcome
    data_source.outcome

## IV. Harmonize Data

Default : automatically tries to infer forward strand alleles using allele frequency (action = 2)
To drop palindromic SNPs, set action = 3

Basic
```{r}
dat <- harmonise_data(
    exposure_dat = bmi_exp_dat,
    outcome_dat = chd_out_dat
)
```

Drop duplicate GWAS summary sets to reduce duplicate MR analyses

PRUNE DATASETS SO ONLY E-O COMBINATIONS WITH HIGHEST POWER/SAMPLE SIZE ARE KEPT FOR BINARY OUTCOMES.


```{r}
dat1<-power_prune(dat,method=1,dist.outcome="binary")
```

PRUNE DATASETS SO ONLY E-O COMBOS WITH HIGHEST INSTRUMENT STRENGTH(LOTS OF VARIATION EXPLAINED) AND SAMPLE SIZE

helpful for studies if there are large number of SNPs available to instrument an exposure. Power can be increased if we prune based on the outcome GWAS with better SNP coverage
```{r}
dat2<-power_prune(dat,method=2,dist.outcome="binary")
```


assumes SNP-exposure effects are continuous with ~N distribution
SNP -outcome effects should be binary or continuous, if binary, use method 1.


### V. MR Analyses


