---
title: "Additional Biomarkers"
author: "Christine Le"
date: "1/18/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preparation

## Load Libraries

```{r}
library(TwoSampleMR)
library(MRInstruments)
library(tidyverse) #so we can use purrr::map, and no, not just because I got cats
library(knitr)
library(ieugwasr)
library(arsenal)
library(tidyverse)
library(multtest)
```

## Data
```{r}
data("gwas_catalog")
plink <- genetics.binaRies::get_plink_binary()

# bfile is the filepath were your reference datasets are located
#no need to include extensions for population (e.g EUR.bin)
bfile <- "data/LD/LD_reference/EUR"

```

# Telomere Length
found in gwas catalog

# Insulin-like growth factor 1/IGF-1


## obtain exposure data

```{r}
#list of exposures
igf1_exp<- gwas_catalog %>% 
  filter(MAPPED_TRAIT_EFO == "IGF-1 measurement") %>% 
  format_data() %>%  
  rename(rsid = SNP)%>%
  clump_data()


```

## Cases 90

```{r}
# obtain outcome data
outcome <- read_outcome_data(
    snps = igf1_exp$SNP,
    filename = "data/outcomes/cases90rsid.csv",
    sep = ",",
    snp_col = "rsid",
    beta_col = "Beta",
    se_col = "SE",
    effect_allele_col = "EA",
    other_allele_col = "NEA",
    eaf_col = "EAF",
    pval_col = "pval",
    samplesize_col = "effective_N")

#harmonize

harm_igf1_cases90 <- harmonise_data(igf1_exp, outcome)
mr_igf1_cases90<- mr(harm_igf1_cases90)

```
## Cases 99

```{r}
# obtain outcome data
outcome <- read_outcome_data(
    snps = igf1_exp$SNP,
    filename = "data/outcomes/cases99rsid.csv",
    sep = ",",
    snp_col = "rsid",
    beta_col = "Beta",
    se_col = "SE",
    effect_allele_col = "EA",
    other_allele_col = "NEA",
    eaf_col = "EAF",
    pval_col = "pval",
    samplesize_col = "effective_N")

#harmonize

harm_igf1_cases99 <- harmonise_data(igf1_exp, outcome)
mr_igf1_cases99<- mr(harm_igf1_cases99)

```

## Pilling

```{r}
# obtain outcome data
outcome <- read_outcome_data(
    snps = igf1_exp$SNP,
    filename = "data/outcomes/29227965-GCST006698-EFO_0007796-build37.f.tsv",
    sep = "\t",
    snp_col = "variant_id",
    beta_col = "beta",
    se_col = "standard_error",
    effect_allele_col = "effect_allele",
    other_allele_col = "other_allele",
    eaf_col = "effect_allele_frequency",
    pval_col = "p_value")

#harmonize

harm_igf1_pilling <- harmonise_data(igf1_exp, outcome)

mr_igf1_pilling<- mr(harm_igf1_pilling)
```


## Timmers

```{r}
# obtain outcome data
outcome <- read_outcome_data(
    snps = igf1_exp$SNP,
    filename = "data/outcomes/lifegen_phase2_bothpl_alldr_2017_09_18.tsv",
    sep = "\t",
    snp_col = "rsid",
    beta_col = "beta1",
    se_col = "se",
    effect_allele_col = "a1",
    other_allele_col = "a0",
    eaf_col = "freq1",
    pval_col = "p")

#harmonize

harm_igf1_timmers <- harmonise_data(igf1_exp, outcome)
mr_igf1_timmers<- mr(harm_igf1_timmers)

```

# IGFBP3 (insulin-like growth factor binding protein 3)
 
## obtain exposure data

```{r}
#list of exposures
igfbp3_exp_ls <- c("IGFBP-3", "Insulin-like growth factor-binding protein 3")

igfbp3_gwas <- available_gwas %>%
  filter(trait %in% igfbp3_exp_ls)%>% 
  pull(id)

#extract exposure data

igfbp3_exp <- extract_instruments(igfbp3_gwas )


```

## Cases 90

```{r}
# obtain outcome data
outcome <- read_outcome_data(
    snps = igfbp3_exp$SNP,
    filename = "data/outcomes/cases90rsid.csv",
    sep = ",",
    snp_col = "rsid",
    beta_col = "Beta",
    se_col = "SE",
    effect_allele_col = "EA",
    other_allele_col = "NEA",
    eaf_col = "EAF",
    pval_col = "pval",
    samplesize_col = "effective_N")

#harmonize

harm_igfbp3_cases90 <- harmonise_data(igfbp3_exp, outcome)
mr_igfbp3_cases90<- mr(harm_igfbp3_cases90)

```
## Cases 99

```{r}
# obtain outcome data
outcome <- read_outcome_data(
    snps = igfbp3_exp$SNP,
    filename = "data/outcomes/cases99rsid.csv",
    sep = ",",
    snp_col = "rsid",
    beta_col = "Beta",
    se_col = "SE",
    effect_allele_col = "EA",
    other_allele_col = "NEA",
    eaf_col = "EAF",
    pval_col = "pval",
    samplesize_col = "effective_N")

#harmonize

harm_igfbp3_cases99 <- harmonise_data(igfbp3_exp, outcome)
mr_igfbp3_cases99 <- mr(harm_igfbp3_cases99)

```

## Pilling

```{r}
# obtain outcome data
outcome <- read_outcome_data(
    snps = igfbp3_exp$SNP,
    filename = "data/outcomes/29227965-GCST006698-EFO_0007796-build37.f.tsv",
    sep = "\t",
    snp_col = "variant_id",
    beta_col = "beta",
    se_col = "standard_error",
    effect_allele_col = "effect_allele",
    other_allele_col = "other_allele",
    eaf_col = "effect_allele_frequency",
    pval_col = "p_value")

#harmonize

harm_igfbp3_pilling <- harmonise_data(igfbp3_exp, outcome)
mr_igfbp3_pilling <- mr(harm_igfbp3_pilling)

```


## Timmers

```{r}
# obtain outcome data
outcome <- read_outcome_data(
    snps = igfbp3_exp$SNP,
    filename = "data/outcomes/lifegen_phase2_bothpl_alldr_2017_09_18.tsv",
    sep = "\t",
    snp_col = "rsid",
    beta_col = "beta1",
    se_col = "se",
    effect_allele_col = "a1",
    other_allele_col = "a0",
    eaf_col = "freq1",
    pval_col = "p")

#harmonize

harm_igfbp3_timmers <- harmonise_data(igfbp3_exp, outcome)

mr_igfbp3_timmers <- mr(harm_igfbp3_timmers)
```

# IL-6
found in gwas catalog

# CRP
 
## obtain exposure data

```{r}
#list of exposures
crp_exp_ls <- c("CRP", "C-reactive protein", "C-Reactive protein level", "C-reactive protein levels")

crp_gwas <- available_gwas %>%
  filter(trait %in% crp_exp_ls)%>% 
  pull(id)

#extract exposure data

crp_exp <- extract_instruments(crp_gwas)


```

## Cases 90

```{r}
# obtain outcome data
outcome <- read_outcome_data(
    snps = crp_exp$SNP,
    filename = "data/outcomes/cases90rsid.csv",
    sep = ",",
    snp_col = "rsid",
    beta_col = "Beta",
    se_col = "SE",
    effect_allele_col = "EA",
    other_allele_col = "NEA",
    eaf_col = "EAF",
    pval_col = "pval",
    samplesize_col = "effective_N")

#harmonize

harm_crp_cases90 <- harmonise_data(crp_exp, outcome)

mr_crp_cases90 <- mr(harm_crp_cases90)
```
## Cases 99

```{r}
# obtain outcome data
outcome <- read_outcome_data(
    snps = crp_exp$SNP,
    filename = "data/outcomes/cases99rsid.csv",
    sep = ",",
    snp_col = "rsid",
    beta_col = "Beta",
    se_col = "SE",
    effect_allele_col = "EA",
    other_allele_col = "NEA",
    eaf_col = "EAF",
    pval_col = "pval",
    samplesize_col = "effective_N")

#harmonize

harm_crp_cases99 <- harmonise_data(crp_exp, outcome)
mr_crp_cases99 <- mr(harm_crp_cases99)

```

## Pilling

```{r}
# obtain outcome data
outcome <- read_outcome_data(
    snps = crp_exp$SNP,
    filename = "data/outcomes/29227965-GCST006698-EFO_0007796-build37.f.tsv",
    sep = "\t",
    snp_col = "variant_id",
    beta_col = "beta",
    se_col = "standard_error",
    effect_allele_col = "effect_allele",
    other_allele_col = "other_allele",
    eaf_col = "effect_allele_frequency",
    pval_col = "p_value")

#harmonize

harm_crp_pilling <- harmonise_data(crp_exp, outcome)

mr_crp_pilling <- mr(harm_crp_pilling)
```


## Timmers

```{r}
# obtain outcome data
outcome <- read_outcome_data(
    snps = crp_exp$SNP,
    filename = "data/outcomes/lifegen_phase2_bothpl_alldr_2017_09_18.tsv",
    sep = "\t",
    snp_col = "rsid",
    beta_col = "beta1",
    se_col = "se",
    effect_allele_col = "a1",
    other_allele_col = "a0",
    eaf_col = "freq1",
    pval_col = "p")

#harmonize

harm_crp_timmers <- harmonise_data(crp_exp, outcome)
mr_crp_timmers <- mr(harm_crp_timmers)

```


# TNF-a
found in gwas catalog under obesity traits

# Cystatin C

found in gwas catalog

## obtain exposure data

```{r}
#list of exposures
crp_exp_ls <- c("CRP", "C-reactive protein", "C-Reactive protein level", "C-reactive protein levels")

crp_gwas <- available_gwas %>%
  filter(trait %in% crp_exp_ls)%>% 
  pull(id)

#extract exposure data

crp_exp <- extract_instruments(crp_gwas)


```

## Cases 90

```{r}
# obtain outcome data
outcome <- read_outcome_data(
    snps = crp_exp$SNP,
    filename = "data/outcomes/cases90rsid.csv",
    sep = ",",
    snp_col = "rsid",
    beta_col = "Beta",
    se_col = "SE",
    effect_allele_col = "EA",
    other_allele_col = "NEA",
    eaf_col = "EAF",
    pval_col = "pval",
    samplesize_col = "effective_N")

#harmonize

harm_crp_cases90 <- harmonise_data(crp_exp, outcome)


```
## Cases 99

```{r}
# obtain outcome data
outcome <- read_outcome_data(
    snps = crp_exp$SNP,
    filename = "data/outcomes/cases99rsid.csv",
    sep = ",",
    snp_col = "rsid",
    beta_col = "Beta",
    se_col = "SE",
    effect_allele_col = "EA",
    other_allele_col = "NEA",
    eaf_col = "EAF",
    pval_col = "pval",
    samplesize_col = "effective_N")

#harmonize

harm_crp_cases99 <- harmonise_data(crp_exp, outcome)


```

## Pilling

```{r}
# obtain outcome data
outcome <- read_outcome_data(
    snps = crp_exp$SNP,
    filename = "data/outcomes/29227965-GCST006698-EFO_0007796-build37.f.tsv",
    sep = "\t",
    snp_col = "variant_id",
    beta_col = "beta",
    se_col = "standard_error",
    effect_allele_col = "effect_allele",
    other_allele_col = "other_allele",
    eaf_col = "effect_allele_frequency",
    pval_col = "p_value")

#harmonize

harm_crp_pilling <- harmonise_data(crp_exp, outcome)


```


## Timmers

```{r}
# obtain outcome data
outcome <- read_outcome_data(
    snps = crp_exp$SNP,
    filename = "data/outcomes/lifegen_phase2_bothpl_alldr_2017_09_18.tsv",
    sep = "\t",
    snp_col = "rsid",
    beta_col = "beta1",
    se_col = "se",
    effect_allele_col = "a1",
    other_allele_col = "a0",
    eaf_col = "freq1",
    pval_col = "p")

#harmonize

harm_crp_timmers <- harmonise_data(crp_exp, outcome)


```

# NT-proBNP
not found
# Insulin
## obtain exposure data

```{r}
#list of exposures
insulin_exp_ls <- c("Insulin")

insulin_gwas <- available_gwas %>%
  filter(trait %in% insulin_exp_ls)%>% 
  pull(id)

#extract exposure data

insulin_exp <- extract_instruments(insulin_gwas )%>% clump_data()




```

## Cases 90

```{r}
# obtain outcome data
outcome <- read_outcome_data(
    snps = insulin_exp$SNP,
    filename = "data/outcomes/cases90rsid.csv",
    sep = ",",
    snp_col = "rsid",
    beta_col = "Beta",
    se_col = "SE",
    effect_allele_col = "EA",
    other_allele_col = "NEA",
    eaf_col = "EAF",
    pval_col = "pval",
    samplesize_col = "effective_N")

#harmonize

harm_insulin_cases90 <- harmonise_data(insulin_exp, outcome)

mr_insulin_cases90 <- mr(harm_insulin_cases90)
```
## Cases 99

```{r}
# obtain outcome data
outcome <- read_outcome_data(
    snps = insulin_exp$SNP,
    filename = "data/outcomes/cases99rsid.csv",
    sep = ",",
    snp_col = "rsid",
    beta_col = "Beta",
    se_col = "SE",
    effect_allele_col = "EA",
    other_allele_col = "NEA",
    eaf_col = "EAF",
    pval_col = "pval",
    samplesize_col = "effective_N")

#harmonize

harm_insulin_cases99 <- harmonise_data(insulin_exp, outcome)

mr_insulin_cases99 <- mr(harm_insulin_cases99)
```

## Pilling

```{r}
# obtain outcome data
outcome <- read_outcome_data(
    snps = insulin_exp$SNP,
    filename = "data/outcomes/29227965-GCST006698-EFO_0007796-build37.f.tsv",
    sep = "\t",
    snp_col = "variant_id",
    beta_col = "beta",
    se_col = "standard_error",
    effect_allele_col = "effect_allele",
    other_allele_col = "other_allele",
    eaf_col = "effect_allele_frequency",
    pval_col = "p_value")

#harmonize

harm_insulin_pilling <- harmonise_data(insulin_exp, outcome)

mr_insulin_pilling <- mr(harm_insulin_pilling)
```


## Timmers

```{r}
# obtain outcome data
outcome <- read_outcome_data(
    snps = insulin_exp$SNP,
    filename = "data/outcomes/lifegen_phase2_bothpl_alldr_2017_09_18.tsv",
    sep = "\t",
    snp_col = "rsid",
    beta_col = "beta1",
    se_col = "se",
    effect_allele_col = "a1",
    other_allele_col = "a0",
    eaf_col = "freq1",
    pval_col = "p")

#harmonize

harm_insulin_timmers <- harmonise_data(insulin_exp, outcome)
mr_insulin_timmers <- mr(harm_insulin_timmers)

```

# GDF15

## obtain exposure data

```{r}
gdf15_exp_ls <- c("growth differentiation factor 15")

gdf15_gwas <- available_gwas %>%
  filter(trait %in% gdf15_exp_ls)%>% 
  pull(id)

#extract exposure data

gdf15_exp <- extract_instruments(gdf15_gwas )




```

## Cases 90

```{r}
# obtain outcome data
outcome <- read_outcome_data(
    snps = gdf15_exp$SNP,
    filename = "data/outcomes/cases90rsid.csv",
    sep = ",",
    snp_col = "rsid",
    beta_col = "Beta",
    se_col = "SE",
    effect_allele_col = "EA",
    other_allele_col = "NEA",
    eaf_col = "EAF",
    pval_col = "pval",
    samplesize_col = "effective_N")

#harmonize

harm_gdf15_cases90 <- harmonise_data(gdf15_exp, outcome)

mr_gdf15_cases90 <- mr(harm_gdf15_cases90)
```
## Cases 99

```{r}
# obtain outcome data
outcome <- read_outcome_data(
    snps = gdf15_exp$SNP,
    filename = "data/outcomes/cases99rsid.csv",
    sep = ",",
    snp_col = "rsid",
    beta_col = "Beta",
    se_col = "SE",
    effect_allele_col = "EA",
    other_allele_col = "NEA",
    eaf_col = "EAF",
    pval_col = "pval",
    samplesize_col = "effective_N")

#harmonize

harm_gdf15_cases99 <- harmonise_data(gdf15_exp, outcome)

mr_gdf15_cases99 <- mr(harm_gdf15_cases99)
```

## Pilling

```{r}
# obtain outcome data
outcome <- read_outcome_data(
    snps = igf1_exp$SNP,
    filename = "data/outcomes/29227965-GCST006698-EFO_0007796-build37.f.tsv",
    sep = "\t",
    snp_col = "variant_id",
    beta_col = "beta",
    se_col = "standard_error",
    effect_allele_col = "effect_allele",
    other_allele_col = "other_allele",
    eaf_col = "effect_allele_frequency",
    pval_col = "p_value")

#harmonize

harm_gdf15_pilling <- harmonise_data(gdf15_exp, outcome)

mr_gdf15_pilling <- mr(harm_gdf15_pilling)
```


## Timmers

```{r}
# obtain outcome data
outcome <- read_outcome_data(
    snps = igf1_exp$SNP,
    filename = "data/outcomes/lifegen_phase2_bothpl_alldr_2017_09_18.tsv",
    sep = "\t",
    snp_col = "rsid",
    beta_col = "beta1",
    se_col = "se",
    effect_allele_col = "a1",
    other_allele_col = "a0",
    eaf_col = "freq1",
    pval_col = "p")

#harmonize

harm_gdf15_timmers <- harmonise_data(gdf15_exp, outcome)

mr_gdf15_timmers <- mr(harm_gdf15_timmers)
```

# IGFBPs
in gwas catalog


# Supplementary table 2 with IEU ids


# table of MR results